Hydras and Cie \\ Variations on Kirby \& Paris' hydra battles and ordinal numbers in Coq

### 2021-09-24 doc/JFLA2022/paper.tex
Add introducting subsection on the vision of the project.

\begin{theorem}
  In the Hydra game, Hercules eventually wins, whichever the strategy of both players :
  choice of a head to chop off, choice of the number of copies. 
 \label{kp:thm1}
\end{theorem}

@@

 In mathematical terms, if at step $n$ the hydra is associated with the ordinal $\alpha$, at step $n+1$ it is associated with
 $\canonseq{\alpha}{n+1}$, the $(n+1)$th element of the canonical sequence of $\alpha$~\cite{KS81}.

@@

We also study the number of steps of a battle:
Let $\alpha<\epsilon_0$ be an ordinal.
Using the \texttt{coq-equations} plug-in~\cite{sozeau:hal-01671777}, we define the function which returns the number of steps of the battle starting with $\alpha$ at step $i$, for  any $i$ in $\mathbb{N}$.
We prove that this number is greater or equal than
$H'_\alpha(i)-i$, where $H'$ is a slight variant of the Hardy hierarchy of rapidly growing functions~\cite{BW85, KS81, Promel2013, Wainer1970}, defined by transfinite recursion in Fig.~\ref{fig:Hprime}. 
From this result, we infer that the considered function is not primitive recursive in general (for $\alpha\geq\omega^\omega$).

@@

Computability is a key topic in Computer Science teaching. Moreover O'Connor's library is a nice illustration of dependently typed programming, so we chose to devote a full chapter to this formalisation, with comments on the definitions and proofs, [counter]-examples and exercises.

@@

  \item Presentation of the main steps of proofs, possibly interleaved with text.
  \item \TODO[Clément]{A good part of the LaTeX support in Alectryon was developed for hydra-battles, and this was the first real application with a custom driver, too.}
  \item \TODO[Clément]{If you kept a trace of them, I would love to see a discussion of any errors that were caught by including the output systematically instead of copy-pasting it.}
  \item \TODO[Pierre]{the order of appearance of snippets in the book is independent from the module structure}
   \item \TODO[Pierre]{Slicing our proofs into snippets allowed to show only the main parts of our proofs (Figure~\ref{fig:Ex42-schutte}).}

@@

The Gaia project~\cite{Gaia}, also maintained on \textit{Coq-Community} contains a development by José Grimm~\cite{grimm:hal-00911710}. This library -- dedicated  to the implementation in \textit{Coq (SSreflect)} of books from  Bourbaki's Elements of Mathematics --- contains a generalization of two types of ordinals (\texttt{T1} for the ordinal $\epsilon_0$ and
\texttt{T2} for $\Gamma_0$) to a larger ordinal defined by Ackermann, and relates these data structures with Bourbaki's set theory.
It is interesting to note that both libraries \textit{Gaia} and \textit{hydra-battles} contain their own (isomorphic but not identical) definitions of types \texttt{T1} and \texttt{T2}, both originated in~\cite{CantorContrib}.
Our plan is to build a  bridge between  the combinatorial results of \textit{hydra-battles} and the set-theoretic content of \textit{Gaia}, and make possible the transfer of theorems from one of the two libraries to the other one.

>>>>

\TODO[Clément]{We haven't defined winning strategies}
\begin{theorem}
  In the Hydra game, every strategy is a winning strategy for Hercules. \label{kp:thm1}
\end{theorem}

@@

 It mathematical terms, if at step $n$ the hydra is associated with the ordinal $\alpha$, at step $n+1$ it is associated with
 $\canonseq{\alpha}{n+1}$, the $(n+1)$th element of the canonical sequence of $\alpha$~\cite{KS81}.

@@

We also study the number of steps of a battle:
Let $\alpha<\epsilon_0$ be an ordinal.
Using the \texttt{coq-equations} plug-in~\cite{sozeau:hal-01671777}, we define the function which returns the number of steps of the battle starting with $\alpha$ at step $i$, for  any $i$ in $\mathbb{N}$.
We prove that this number is greater or equal than
$H'_\alpha(i)-i$, where $H'$ is a slight variant of the Wainer Hardy hierarchy of rapidly growing functions~\cite{BW85, KS81, Promel2013, Wainer1970}, defined by transfinite recursion in Fig.~\ref{fig:Hprime}. \TODO[Clément]{It seems that the term ``Wainer Hardy'' hierarchy only appears in his paper and the Hydra-battles book}
From this result, we infer that the considered function is not primitive recursive in general (for $\alpha\geq\omega^\omega$).

@@

Computability is a key topic in Computer Science teaching. Moreover O'Conor's library is a nice illustration of dependently typed programming, so we chose to devote a full chapter to this formalisation, with comments on the definitions and proofs, [counter]-examples and exercises.

@@

  \item Presentation of the main steps of proofs, possibly interleaved with text.
  \item \TODO[Clément]{A good part of the LaTeX support in Alectryon was developed for hydra-battles, and this was the first real application with a custom driver, too.}
  \item \TODO[Clément]{If you kept a trace of them, I would love to see a discussion of any errors that were caught by including the output systematically instead of copy-pasting it.}

@@

The Gaia project~\cite{Gaia}, also maintained on \textit{Coq-Community} contains a development by José Grimm~\cite{grimm:hal-00911710}. This library -- dedicated  to the implementation in \textit{Coq (SSreflect)} of books from  Bourbaki's Elements of Mathematics --- contains a generalization of two types of ordinals (\texttt{T1} for the ordinal $\epsilon_0$ and
\texttt{T2} for $\Gamma_0$) to a larger ordinal defined by Ackermann, and relates these data structures with Bourbaki's set theory.
It is interesting to note that both libraries \textit{Gaia} and \textit{hydra-battles} contain their own (isomorphic but not identical) definitions of types \texttt{T1} and \texttt{T2}, both originated in~\cite{CantorContrib}.
Our plan is to build a  ``gate'' \TODO[Clément]{Bridge?} giving access to both the combinatorial results of \textit{hydra-battles} and the set-theoretic content of \textit{Gaia}, and make possible the transfer of theorems from one of the two libraries to the other one.
